---
title: "dv_final-project"
author: "Misha Ash"
date: "5/28/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


[final project requirements](https://github.com/uc-dataviz/course/blob/master/assignments/final-project.md)

### presentation tasks

- [x] use [flexdashboard storyboard](https://rmarkdown.rstudio.com/flexdashboard/using.html#storyboards)

- [ ] use [relative paths!](https://gist.github.com/jennybc/362f52446fe1ebc4c49f) 



### presentation outline

- brief description of the study and design
  - structure of trials: *dendrogram* (ggdendro)
- descriptive stats
  - mosaic charts?
  - waffle charts?
  - 
- results
  - typical presentation of mean differences
  - downsides
    - Piccinini's criticism of bar graphs
  - adding data points (possible solution)
    - problem: uninformative because of overlap
    - [example](http://flowingdata.com/2015/08/11/real-chart-rules-to-follow/) of showing data
  - violin plots (alternative)


### final project outline

[ ] use R markdown html output for final report
[ ] focus on DV principles
[ ] if time, consider representing conditions with Venn diagram


* when to show the distribution
* distributions and mean differences
* the role of time
* time and mean differences



### dataset encodings for final project

```{r level 1 dataset encoding}

study_01 <- readr::read_csv("~/R/CGChains/data/CGChains_CODED-ELAN-DATA_2018-04-24.csv")

# no interaction terms
level_01_data <- study_01 %>%

# INDEX VARs
  
  # ITEM: individual stimulus problem numbers
  mutate(ITEM = as.integer(ITEM_CODE)) %>%
  # LIST
  mutate(LIST = LIST) %>%
  # TRIAL
  mutate(TRIAL = trial_number) %>%
  # SUBJECT ID
  mutate(ID = SUBJ) %>%

# PRIMARY IVs

  # SPATIALITY: spatial trials = 1, non-spatial trials = 0
  mutate(SPACE_IV = if_else(study_01$trial_type == "spatial", 1, 0)) %>%
  # AXIS: vertical trials = 1, horizontal trials = 0
  mutate(AXIS_IV = if_else(study_01$item_axis == "vertical", 1, 0)) %>%
  # DIFFICULTY: difficult trials = 2, easy trials = 1, invalid trials = 0
  mutate(DIFF_IV = if_else(study_01$DIFFICULTY_with_invalid == "HARD", 2,
                      if_else(study_01$DIFFICULTY_with_invalid == "EASY", 1, 0))) %>%

  # ORDER: spatial first = 1, non-spatial first = 0
  mutate(ORDER = as.integer(if_else(study_01$SPATIAL_FIRST == 1, 1, 0))) %>%
  
# SECONDARY IVs ## leaving this out for now
  # difficulty based on deSoto (1965): HARDEST = 4, HARD = 3, EASY = 2, EASIEST = 1
  #mutate(deSoto_difficulty_IV2 = cut(x = correct_deSoto_without_invalid, breaks = c(38.3, 41.5, 60.5, 61.8, 100),
                                 #right = FALSE, labels = c(4, 3, 2, 1))) %>%
  
# DVs
  # GESTURE_PRESENCE: present = 1, absent = 0
  mutate(GESTURE_DV = if_else(study_01$gesture_presence == "yes", 1, 0)) %>%
  # GESTURE_AXIS: lateral = 1, vertical = 2, saggital = 3, other/unclear = 4, ERROR = 5
  mutate(AXIS_DV = if_else(study_01$gesture_axis == "lateral", 1, 
                   if_else(study_01$gesture_axis == "vertical", 2,
                   if_else(study_01$gesture_axis == "sagittal", 3, 
                   if_else(study_01$gesture_axis == "other/unclear", 4, 5))))) %>%

  # ACCURACY: correct = 1, incorrect = 0
  mutate(ACCURACY_DV = if_else(study_01$ACCURACY == 1, 1, 0)) %>%
  
# SELECT VARIABLES

  dplyr::select(ITEM, LIST, TRIAL, ID, 
         SPACE_IV, AXIS_IV, DIFF_IV, 
         ORDER, # deSoto_difficulty_IV2, 
         GESTURE_DV, AXIS_DV, ACCURACY_DV)
```

```{r gesture & accuracy by subject}
# calculate gesture and accuracy proportions by subject
X <- level_01_data %>%
  #dplyr::select(SUBJ, GESTURE_DV, ACCURACY_DV) %>%

  dplyr::group_by(ID, SPACE_IV) %>%
  dplyr::summarize(gestures = sum(GESTURE_DV), 
                   percent_gestured = gestures / (length(unique(ITEM))*2),
                   gesture_se = sd(GESTURE_DV)/sqrt((length(unique(ITEM)))*2),
                   
                   errors = length(unique(ITEM)) - sum(ACCURACY_DV),
                   percent_errors = errors / (length(unique(ITEM))*2),
                   error_se = sd(1 - ACCURACY_DV)/sqrt((length(unique(ITEM)))*2)
                   ) %>%

  as_factor(ID) #%>%
  
#    dplyr::arrange(ID) %>%
    # add variable to relabel x axis sequentially
 #  dplyr::mutate(ID_gestures_ascending = as.integer(seq(1:length(unique(study_01$SUBJ))))) #%>%


 
  # order by proportion
  dplyr::arrange(percent_gestured) #%>%
  # add se variables for gesture production and error
  dplyr::mutate(percent_gestured = gestures / length(unique(study_01$ITEM_CODE)),
                percent_errors = errors / length(unique(study_01$ITEM_CODE))) %>%
  
  
  
 

  
  dplyr::arrange(percent_errors) %>%
  
  dplyr::mutate(ID_errors_ascending = as.integer(seq(1:length(unique(study_01$SUBJ))))) %>%
  # could not get color to work without adding a constant
  dplyr::mutate(fill_constant = 1)


summary_subj_err_gest <- mean_subject_errors_gestures %>%
  dplyr::summarize(mean_gestures = mean(gestures), 
                   mean_percent_gestured = mean(percent_gestured),
                   mean_gesture_se = mean(gesture_se),
                   
                   mean_errors = mean(errors),
                   mean_percent_errors = mean(percent_errors),
                   mean_error_se = mean(error_se))
```


### sketches for final project

#### boxplots

```{r box plot}



```


#### distributions

##### static histogram

To begin, I used ggplotly to add interactive plotly labels to some pretty extensively formatted ggplot objects.

Gesture production (i.e., whether a participant gestured on a given problem) is an important response variable in my data. A distribution of this variable may be displayed as a histogram showing the proportion of trials on which each participant gestured:

```{r gesture production by participant, include = FALSE}

# COLORS: using potion brewer color palette from color-hex.com

mean_subject_errors_gestures %>%
  
  ggplot() +
  
  geom_col(aes(x = ID_gestures_ascending, y = percent_gestured, #group = SUBJ, 
                 fill = as.factor(fill_constant), alpha = .75)) +
  
    # change color scheme of fill colors
    scale_fill_manual("fill_constant", values ="#a2adba") +
  
    # add horizontal line showing mean
    geom_hline(aes(yintercept = mean(percent_gestured)), color = "#919ba7") +
  
    # add label with mean value
    geom_text(x = 4, y = 0.56, aes(label = "mean = 0.524"), color = "#919ba7") +
  
    scale_y_continuous(expand = c(0, 0)) +
    scale_x_continuous(breaks = c(seq(1:24)), expand = c(0, 0)) +
  
    theme_light() +
    theme(panel.grid.major.x = element_blank(), # remove vertical grid lines
        panel.grid.minor.x = element_blank(), # remove vertical grid lines
        panel.grid.minor.y = element_blank(), # remove minor horizontal grid lines
        legend.position = "none", # remove legends
        panel.border = element_blank(), # remove outer panel border
        axis.ticks.y = element_blank(), # remove ticks from y axis
        plot.title = element_text(margin = margin(t = 0, r = 0, b = 30, l = 0)),
        axis.title.y = element_text(margin = margin(t = 0, r = 15, b = 0, l = 0)), 
        axis.title.x = element_text(margin = margin(t = 15, r = 0, b = 0, l = 0)), 
        text = element_text(family = "Arial")) +
  
    ggtitle('participant gesture production') +
    xlab('participant (arranged by ascending gesture production)') +
    ylab('proportion of trials with gesture')
```






DESIGN CHOICES

* removed y tick marks


##### interactive histogram



```{r distribution of gesture production}

'total mean =' <- round(mean(X$percent_gestured)*2, digits = 3)


dist_gest <- X %>%
  
  ggplot() +
    
  # mean gesture production line
#  geom_hline(aes(yintercept = mean(percent_gestured)*2), color = "#2e184b", size = 0.3) +
  
      geom_hline(aes(yintercept = mean(percent_gestured)*2, 'total mean' = 0.524), 
              color = "#2e184b", size = 0.3) +

## OLD VERSION  
##    geom_hline(aes(yintercept = mean(percent_gestured)*2, 'total mean' = paste("total mean", mean(percent_gestured)*2)), 
##                 color = "#2e184b", size = 0.3) +
  
  # label mean line
##  geom_text(x = 3, y = (mean(X$percent_gestured)*2 + 0.03), 
##            aes(label = "total mean = "), color = "#2e184b") +
##  
##  geom_text(x = 6, y = (mean(X$percent_gestured)*2 + 0.03), 
##            aes(label = round(mean(percent_gestured)*2, digits = 3)), color = "#2e184b") +
  
  ### REFERENCE
    
   #     geom_text(aes(label = sprintf("%0.2f", round(prop_gesture, digits = 2)), 
    #                y = prop_gesture + 0.1,
     #               group = item_axis,
      #              fontface = "bold"), 
       #             position = position_dodge(0.55),
        #            color = "#767676") +
  
### group means... cluncky code: there must be a simpler way!
  
  # spatial mean line
  geom_hline(aes(yintercept = 
      (sum(X[X$SPACE_IV == 1, "percent_gestured"]) / 
      tally(X[X$SPACE_IV == 1, "percent_gestured"]))[1,1], 'spatial mean' = "X"),
      color = "#5bae94") +
  
  # spatial mean labels  
  geom_text(x = 3, 
            y = (sum(X[X$SPACE_IV == 1, "percent_gestured"]) / 
              tally(X[X$SPACE_IV == 1, "percent_gestured"])[1,1] + 0.03)[1,1], 
            aes(label = "spatial mean = "), color = "#5bae94") +
  
  geom_text(x = 7, 
            y = (sum(X[X$SPACE_IV == 1, "percent_gestured"]) / 
              tally(X[X$SPACE_IV == 1, "percent_gestured"])[1,1] + 0.03)[1,1], 
            
            aes(label = round((sum(X[X$SPACE_IV == 1, "percent_gestured"]) / 
      tally(X[X$SPACE_IV == 1, "percent_gestured"]))[1,1], digits = 3)), color = "#5bae94") +
  
  # non-spatial mean labels  
  geom_text(x = 3.5, 
            y = (sum(X[X$SPACE_IV == 0, "percent_gestured"]) / 
              tally(X[X$SPACE_IV == 0, "percent_gestured"])[1,1] + 0.03)[1,1], 
            aes(label = "non-spatial mean = "), color = "#5bae94") +
  
  geom_text(x = 7.5, 
            y = (sum(X[X$SPACE_IV == 0, "percent_gestured"]) / 
              tally(X[X$SPACE_IV == 0, "percent_gestured"])[1,1] + 0.03)[1,1], 
            
            aes(label = round((sum(X[X$SPACE_IV == 0, "percent_gestured"]) / 
      tally(X[X$SPACE_IV == 0, "percent_gestured"]))[1,1], digits = 3)), color = "#5bae94") +
  
  
#  geom_text(x = 6, y = 0.554, aes(label = round(mean(percent_gestured)*2, digits = 3)), 
#                                  color = "#2e184b") +
  
    # non-spatial mean
  geom_hline(aes(yintercept = 
      (sum(X[X$SPACE_IV == 0, "percent_gestured"]) / 
      tally(X[X$SPACE_IV == 0, "percent_gestured"]))[1,1]), color = "#7e90b6") +
  
#  geom_hline(aes(yintercept = mean((X[X$SPACE_IV == 0, "percent_gestured"])[,])), color = "#FFA319") +
#  geom_hline(aes(yintercept = mean((X[X$SPACE_IV == 1, "percent_gestured"])[,])), color = "#FFA319") +
  

### temporarily remove ribbon (need to fix error margin)
 #   annotate('ribbon', x = c(-Inf, Inf), 
  #           ymin = summary_subj_err_gest$mean_percent_gestured - summary_subj_err_gest$mean_gesture_se, 
   #          ymax = summary_subj_err_gest$mean_percent_gestured + summary_subj_err_gest$mean_gesture_se, 
    #         alpha = 0.2, fill = "#628B9F") +
  
  geom_col(aes(x = reorder(ID, percent_gestured), y = percent_gestured, #group = SUBJ, 
                 fill = as.factor(SPACE_IV), alpha = .85)) +
  

  
   # scale_y_continuous(expand = c(0, 0)) +
   # scale_x_continuous(breaks = c(seq(1:24)), expand = c(0, 0)) +
  
    # change color scheme of fill colors
    scale_fill_manual("fill_constant", values = c("#8da0cb", "#66c2a5")) +
  
  scale_y_continuous(expand = c(0, 0)) +
   # scale_x_continuous(breaks = c(seq(1:24)), expand = c(0, 0)) +
  
    theme_light() +
    theme(panel.grid.major.x = element_blank(), # remove vertical grid lines
        panel.grid.minor.x = element_blank(), # remove vertical grid lines
        panel.grid.minor.y = element_blank(), # remove minor horizontal grid lines
        # remove legends (unnecessary here)
        legend.position = "right", ### TEMP
        plot.title = element_text(margin = margin(t = 0, r = 0, b = 10, l = 0)),
        axis.title.y = element_text(margin = margin(t = 0, r = 15, b = 0, l = 0)), 
        axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0)), 
        panel.border = element_blank(), # remove outer panel border
        axis.ticks.y = element_blank(), # remove ticks from y axis
        text = element_text(family = "Arial")) +
  
    ggtitle('gesture production by participant') +
    xlab('participant id (arranged by total gesture production)') +
    ylab('proportion of trials with gesture')

dist_gest

ggplotly(dist_gest, tooltip = c("total mean", "spatial mean", "all"))




ggsave("gesture_production_dist.png", width = 5.3125, height = 3.270833)
```

