---
title: "dv_final-project"
author: "Misha Ash"
date: "5/28/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


[final project requirements](https://github.com/uc-dataviz/course/blob/master/assignments/final-project.md)

### presentation tasks

- [x] use [flexdashboard storyboard](https://rmarkdown.rstudio.com/flexdashboard/using.html#storyboards)

- [ ] use [relative paths!](https://gist.github.com/jennybc/362f52446fe1ebc4c49f) 



### presentation outline

- brief description of the study and design
  - structure of trials: *dendrogram* (ggdendro)
- descriptive stats
  - mosaic charts?
  - waffle charts?
  - 
- results
  - typical presentation of mean differences
  - downsides
    - Piccinini's criticism of bar graphs
  - adding data points (possible solution)
    - problem: uninformative because of overlap
    - [example](http://flowingdata.com/2015/08/11/real-chart-rules-to-follow/) of showing data
  - violin plots (alternative)


### final project outline

### dataset encodings for final project

```{r level 1 dataset encoding}
# no interaction terms
level_01_data <- study_01 %>%

# INDEX VARs
  
  # ITEM: individual stimulus problem numbers
  mutate(ITEM = as.integer(ITEM_CODE)) %>%
  # LIST
  mutate(LIST = LIST) %>%
  # TRIAL
  mutate(TRIAL = trial_number) %>%
  # SUBJECT ID
  mutate(ID = SUBJ) %>%

# PRIMARY IVs

  # SPATIALITY: spatial trials = 1, non-spatial trials = 0
  mutate(SPACE_IV = if_else(study_01$trial_type == "spatial", 1, 0)) %>%
  # AXIS: vertical trials = 1, horizontal trials = 0
  mutate(AXIS_IV = if_else(study_01$item_axis == "vertical", 1, 0)) %>%
  # DIFFICULTY: difficult trials = 2, easy trials = 1, invalid trials = 0
  mutate(DIFF_IV = if_else(study_01$DIFFICULTY_with_invalid == "HARD", 2,
                      if_else(study_01$DIFFICULTY_with_invalid == "EASY", 1, 0))) %>%

  # ORDER: spatial first = 1, non-spatial first = 0
  mutate(ORDER = as.integer(if_else(study_01$SPATIAL_FIRST == 1, 1, 0))) %>%
  
# SECONDARY IVs ## leaving this out for now
  # difficulty based on deSoto (1965): HARDEST = 4, HARD = 3, EASY = 2, EASIEST = 1
  #mutate(deSoto_difficulty_IV2 = cut(x = correct_deSoto_without_invalid, breaks = c(38.3, 41.5, 60.5, 61.8, 100),
                                 #right = FALSE, labels = c(4, 3, 2, 1))) %>%
  
# DVs
  # GESTURE_PRESENCE: present = 1, absent = 0
  mutate(GESTURE_DV = if_else(study_01$gesture_presence == "yes", 1, 0)) %>%
  # GESTURE_AXIS: lateral = 1, vertical = 2, saggital = 3, other/unclear = 4, ERROR = 5
  mutate(AXIS_DV = if_else(study_01$gesture_axis == "lateral", 1, 
                   if_else(study_01$gesture_axis == "vertical", 2,
                   if_else(study_01$gesture_axis == "sagittal", 3, 
                   if_else(study_01$gesture_axis == "other/unclear", 4, 5))))) %>%

  # ACCURACY: correct = 1, incorrect = 0
  mutate(ACCURACY_DV = if_else(study_01$ACCURACY == 1, 1, 0)) %>%
  
# SELECT VARIABLES
  
  select(ITEM, LIST, TRIAL, ID, 
         SPACE_IV, AXIS_IV, DIFF_IV, 
         ORDER, # deSoto_difficulty_IV2, 
         GESTURE_DV, AXIS_DV, ACCURACY_DV)
```



